---
title: "stats506_hw2"
format:
  html:
    embed-resources: true
editor: source
---

github repo link: https://github.com/juliahui/stats506_hw2.git

```{r}
#install.packages("microbenchmark")
```

```{r}
library(tidyverse)
library(roxygen2)
library(schoolmath) #is.positive
library(microbenchmark)
```

Problem 1: Random Walk with Momentum

first step: all four directions have equal prob (1/4), 
for rest of steps: direction depends on previous step

With probability p, the particle moves in the same direction as the previous 
step.

With probability 1−p, it chooses one of the four directions at random, with 
equal probability (1/4)

a. simulation function (i + ii) write function momentum_walk(n, p) that 
simulates this process and returns the particle's final position, sanitize 
input + b (i). implement simulation: write the function momentum_walk1(n, p) 
using a loop (a for loop or a while loop), briefly explain logic of each 
implementation
```{r}
#'Function to simulate random walk with momentum process
#'
#'@param n number of steps
#'@param p value between 0 and 1 that represents the probability of preserving 
#'the previous direction
#'@return vector containing the particle's final position

momentum_walk1 <- function(n, p) {
  #sanitize input
  if (n < 1) {
    stop("n must be greater than or equal to 1")
  }
  if (!is.positive(n) | (n %% 1) != 0) {
    stop("n must be a positive integer ")
  }
  if (p < 0 || p > 1) {
    stop("p must be between 0 and 1 inclusive")
  }
  
  #vector for current position, at starting point initially
  current_pos <- c(0, 0)
  
  #list of possible steps; choice would be added to current_pos to update
  #right, up, left, down
  possible_steps <- list(c(1, 0), c(0, 1), c(-1, 0), c(0, -1))
  
  #first step: all 4 dirs have prob 1/4, choose the first step
    #default is equal probability, included for clarity
  step_choice <- sample(possible_steps, size = 1, 
                        prob = c(0.25, 0.25, 0.25, 0.25))
  #variable for keeping track of what the step's direction is for later
    #contains vector, eg. 1 0 
  step_dir <- step_choice[[1]]
  #current position update; current_pos: vector, step_dir: vector 
    #pulled from list step1_choice 
  current_pos <- current_pos + step_dir
  
  if (n == 1) {
    return(current_pos)
  }

  #loop through the rest of the steps if there's more than one step
    #P(X = 1) = p (event that we go in the same dir), 
    #P(X = 0) = 1-p (event that we re-choose a direction)
  if (n > 1) {
    for (i in 1:(n-1)) {
      #gives 1 or 0
      same_dir <- rbinom(n = 1, size = 1, prob = p)
      #if the bernoulli trial -> 1: take step in same dir
      if (same_dir == 1) {
        current_pos <- current_pos + step_dir
      #if -> 0, go through same steps as step 1 to choose a direction
      } else {
        step_choice <- sample(possible_steps, size = 1, 
                        prob = c(0.25, 0.25, 0.25, 0.25))
        step_dir <- step_choice[[1]]
        current_pos <- current_pos + step_dir
      }
    }
  }
  return(current_pos)
}
```

```{r}
momentum_walk1(10, 0.6)

momentum_walk1(1000, 0.6)
```

In momentum_walk(), I created a list of possible steps to 'choose' or sample 
from for the first step and any steps later on where we need to re-choose the 
direction. After taking the first step, the direction of that step is recorded, 
and we start to loop through steps 2 through n. Since we either keep the same 
direction as the previous step or choose the direction again, it ends up being 
a Bernoulli trial each loop, where X = 1 has a probability of p and X = 0 has 
a probability of 1-p. Depending on the outcome of the random Bernoulli trial, 
we take a step in the same direction or we go through the process of choosing 
from the four directions again. 

b (ii). write the function momentum_walk2(n, p) using one of the apply functions 
(choose an appropriate member of the apply family), briefly explain logic of 
each implementation

```{r}
#'Function to simulate random walk with momentum process using an lapply
#'
#'@param n number of steps
#'@param p value between 0 and 1 that represents the probability of preserving 
#'the previous direction
#'@return vector containing the particle's final position

momentum_walk2 <- function(n, p) {
  #sanitize input
  if (n < 1) {
    stop("n must be greater than or equal to 1")
  }
  if (!is.positive(n) | (n %% 1) != 0) {
    stop("n must be a positive integer ")
  }
  if (p < 0 || p > 1) {
    stop("p must be between 0 and 1 inclusive")
  }
  
  possible_steps <- list(c(1, 0), c(0, 1), c(-1, 0), c(0, -1))
  #contains direction vector of first step, first step's position
  first_step <- sample(possible_steps, size = 1, 
                        prob = c(0.25, 0.25, 0.25, 0.25))[[1]]
  step_dir <- first_step
  
  #if there's only one step, return the first step's position
  if (n == 1) {
    return(first_step)
  }
  
  #if there's more than one step, get rest of steps' direction vectors (eg. 1 0) 
    #use lapply, then add them all to pos after 1st step => final position
  if (n > 1) {
    #idea to use seq_len(): https://stackoverflow.com/questions/58086033/how-to-pass-the-first-argument-to-lapply-so-that-it-represents-a-list-of-rows
  #rest_of_steps: a list containing the directions for steps 2 through n 
  rest_of_steps <- lapply(seq_len(n-1), function(x) {
    same_dir <- rbinom(n = 1, size = 1, prob = p)
    #if we're not keeping the same direction as previous step
    if (same_dir == 0) {
      #choose direction again
      step_choice <- sample(possible_steps, size = 1, 
                        prob = c(0.25, 0.25, 0.25, 0.25))
      #update step direction; seems to be only way to update it outside lapply
      step_dir <<- step_choice[[1]]
    }
    #return direction from this step
    return(step_dir)
  })
  }
  #all step directions, first_step originally vector, make into list since
    #rest_of_steps is a list 
  all_step_dirs <- c(list(first_step), rest_of_steps)
  
  #add all of the direction vectors together to get final position of particle
  #idea to use Reduce: https://stackoverflow.com/questions/49516132/how-to-sum-list-of-list-of-vector-in-r
  final_pos <- Reduce(`+`, all_step_dirs)
  return(final_pos)
}

momentum_walk2(10, 0.6)

momentum_walk2(1000, 0.6)
```

momentum_walk2 uses similar logic to momentum_walk1, only I'm using lapply() 
instead of a for loop for steps 2 through n. Using lapply() meant that the 
variables had to be updated using <<- in order for the update to be in the 
function rather than within lapply(), even though it's not usually recommended 
to do so. Updating the position each iteration seemed a little complicated in 
lapply(), so instead I just returned each step's direction so that the variable 
rest_of_steps was a list of direction vectors, and then added them all together 
at the end.

c. reproducibility
```{r}
set.seed(10)
momentum_walk1(10, 0.6)
momentum_walk1(1000, 0.6)
```

Using set.seed() ensures that we get the same 'random' sequence every time the
code is run. 

d. does momentum matter? consider p = 0, p = 0.5, p = 0.9

i. for each val of p, simulate walk of 1000 steps
```{r}
set.seed(10)

final_location0 <- momentum_walk1(1000, 0)
dist0 <- sqrt(final_location0[1]^2 + final_location0[2]^2)
dist0

final_location0_5 <- momentum_walk1(1000, 0.5)
dist0_5 <- sqrt(final_location0_5[1]^2 + final_location0_5[2]^2)
dist0_5

final_location0_9 <- momentum_walk1(1000, 0.9)
dist0_9 <- sqrt(final_location0_9[1]^2 + final_location0_9[2]^2)
dist0_9
```

ii. Increasing momentum appears to increase the final distance

iii. A higher p value may mean a greater chance the particle will continue 
moving in the same direction as before, and a lesser chance of moving back to 
its previous position (eg. up then back down or right then back to the left), 
which could increase the final distance.

e. (i + ii) Monte Carlo simulation
```{r}
set.seed(10)

#number of steps in each walk
num_steps <- 1000
#number of walks
num_walks <- 1000
#all the p values we want to get avg final distances for 
p_vals <- c(0, 0.25, 0.5, 0.75, 0.9)

#list of final distance averages
avg_final_dists <- lapply(p_vals, function(p) {
  #replicate/repeat getting the distance num_walks amount of times for current p
  final_dists <- replicate(num_walks, expr = {
    final_location <- momentum_walk1(num_steps, p)
    sqrt(final_location[1]^2 + final_location[2]^2)
  })
  #average the final distances for current p
  mean(final_dists)
})
```


iii.
```{r}
results_table <- data.frame(p = p_vals, mean_dist = unlist(avg_final_dists))
results_table
```

iv. As p increases, the average distance the particle travels increases as well. 

f. compare computational performance

Use the microbenchmark package to compare the speed of your two implementations. 
Compare their performance when n = 1000, p = 0.6, and n = 100000, p = 0.6. 
Summarize your results
```{r}
thousand <- microbenchmark(momentum_walk1(1000, 0.6),
               momentum_walk2(1000, 0.6))

hundred_thousand <- microbenchmark(momentum_walk1(100000, 0.6),
               momentum_walk2(100000, 0.6))
```

i. Which implementation is faster? 

My implementation of momentum_walk1 is faster than momentum_walk2. My 
implementation of momentum_walk2 must not be very good since I'd expected the 
other way around.

ii. Does the difference become larger for the larger input?

The difference doesn't really get larger by much for the larger input. The 
average for momentum_walk2 is about 1.33 for 1000 iterations, and 1.34 for 
100,000.

```{r}
(2.164636/1.621824)*100

(223.6310/166.9582)*100
```

iii. Why might the implementations have different execution times?

My momentum_walk2 implementation is probably slower because I use lapply() to 
return a large list, which I add the first step to, creating a larger list, 
whereas in momentum_walk1 I just update a couple variables. So even though I'm 
looping in momentum_walk1, the way I used lapply() makes momentum_walk2 slower. 
Some other factors might also make momentum_walk2 slower, like my use of <<- or 
maybe repeatedly calling a slightly complicated function within lapply(). 


Problem 2 - Bus Passengers Monte Carlo Estimation
```{r}
set.seed(10)

num_days <- 10000

#from 6 to 8 (3 hrs), follows Pois(4) distribution
  #rpois(num_days, etc) -> generates 10000 datasets, three hours -> mult by 3
six_to_eight <- rpois(n = num_days * 3, 
                      lambda = 4)
six_to_eight_mat <- matrix(six_to_eight, 
                           nrow = 3, 
                           ncol = num_days)
six_to_eight_total <- apply(six_to_eight_mat, 2, sum)

#9 (1 hr), N(45, 16), sd = sqrt(16)
nine_peak <- rnorm(n = num_days, 
                   mean = 45, 
                   sd = 4)

#10-3(6 hrs), Pois(15)
ten_to_three <- rpois(n = num_days * 6, 
                      lambda = 15)
ten_to_three_mat <- matrix(ten_to_three, 
                           nrow = 6, 
                           ncol = num_days)
ten_to_three_total <- apply(ten_to_three_mat, 2, sum)

#4 (1 hr), N(55, 25)
four_peak <- rnorm(n = num_days, 
                   mean = 55, 
                   sd = 5)

#5-10 (6 hrs), Pois(9)
five_to_ten <- rpois(n = num_days * 6, 
                     lambda = 9)
five_to_ten_mat <- matrix(five_to_ten, 
                           nrow = 6, 
                           ncol = num_days)
five_to_ten_total <- apply(five_to_ten_mat, 2, sum)

daily_passenger_totals <- six_to_eight_total + nine_peak + ten_to_three_total +
                          four_peak + five_to_ten_total

estimated_daily_total <- mean(daily_passenger_totals)
estimated_daily_total
```

Based on the Monte Carlo simulation, the estimated number of passengers who 
board the bus per day is about 256. 


Problem 3 - Linear Regression
a. Data preparation and de-identification

Remove all variables that could uniquely identify a particular commercial. 
This includes variables such as the brand, title, URLs, YouTube channel, and 
publication information.

After removing these variables, report the dimensions of the resulting dataset.

```{r}
youtube <- read.csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-03-02/youtube.csv"
)
head(youtube)
dim(youtube)
#get column names
names(youtube) 

#de-identification: remove brand, title, urls, yt channel, publication info
youtube_di <- youtube |>
  select(-c(brand, superbowl_ads_dot_com_url, youtube_url, published_at, title,
            channel_title))
head(youtube_di)
dim(youtube_di)
```

The dimension of the de-identified dataset is 247 rows by 19 columns

b. Exploring potential outcomes

Examine distribution + determine which of following categs. is most appropriate:

The variable can be used as is as the outcome in a linear regression model.

The variable should be transformed before being used as the outcome in a linear regression model.

The variable is not appropriate to use as the outcome in a linear regression model.

Briefly justify your decision for each variable.

If you decide that a transformation is needed, perform an appropriate transformation and use the transformed variable for the remainder of the problem.
```{r}
#view_count, like_count, dislike_count, favorite_count, comment_count
```






























